---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
    %\VignetteIndexEntry{Installation}
      %\usepackage[UTF-8]{inputenc}
---

The package moanin was developped to provide a simple and efficient workflow
for time-course gene expression data. It depends on packages from [bioconductor](https://www.bioconductor.org/)
and [CRAN](https://cran.r-project.org/).

- CRAN packages:

    - MASS (>= 1.0.0)
    - viridis
    - ClusterR
    - zoo

- Bioconductor packages:

    - edge (>= 2.4.0)
    - limma
    - edgeR
    - reshape2
    - NMI
    - topGO

To use this package, install it with

```{r eval=FALSE}
# install devtools
install.packages("devtools")
# use devtools to install moanin
devtools::install_github("NelleV/moanin")
```

Next, load the package into your namespace:

```{r}
library(moanin)
```

And you are ready to go!


# Datasets

The package comes with a couple of datasets:

    - shoemaker2015

## Time-course data of mice lung samples exposed to different influenza strains

```{r}
library(dplyr)
library(knitr)
library(kableExtra)
library(moanin)
library(timecoursedata)
data(shoemaker2015)

data = shoemaker2015$data
meta = shoemaker2015$meta
```

Here's the firt 10 lines of the data named `matrix`:

```{r echo=FALSE}
data = data[1:10, ]
meta = meta[1:10, ]
kable(data) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

and the first 10 lines of the metadata `data.frame`.
```{r echo=FALSE}
kable(meta) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

