---
title: "Implementation details"
output: BiocStyle::html_document
vignette: >
  %\VignetteEngine{knitr::knitr}
    %\VignetteIndexEntry{Implementation Details}
      %\usepackage[UTF-8]{inputenc}
---
```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(
  cache=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE,
  echo=FALSE, out.width=".5\\textwidth",
  duplicate.label="allow", fig.pos <- "H", out.extra = "")
```

```{r}
library(moanin)
data(exampleData)
moaninObject <- create_moanin_model(data=testData, meta=testMeta)
```

# Handling contrasts with a generic formula
Assume we have $L$ levels in our grouping variable. 

For a contrast of the form `"K-C"` where "K" and "C" are two of the three levels of the grouping variable (as in our example data, see `?exampleData`) we currently convert this internally (via the internal `is_contrasts` function) to 

```{r}
contr<-getFromNamespace("is_contrasts","moanin")
t(contr("K-C",moaninObject))
```

However, this "contrast" isn't really the underlying contrast, meaning it's not really a linear combination of all the coefficients in the model -- since it's dimension is not of length $p$. We are fitting a function $f_\ell$ for each level, which is defined by a linear combination of spline basis functions. So our set of coefficients are the coefficients of all the basis functions. So we need contrasts of length $p$, where $p$ is the number of coefficients in our design matrix $X$  (`basis_matrix`). 

In particular, we want to take the contrast for the entire function $f_\ell$, i.e. $f_K=f_C$. That means we want all of the coefficients for $f_K$ and $f_C$ to be equal. (Note -- I guess conceivably you could have a different set of constraints that still resulted in $f_{K}=f_{C}$? )
So the above contrasts need to be done *per basis variable*. 

What we are really doing is the following set of constraints that we are imposing on our $p$ coefficients in our design matrix $X$ the linear model simultaneously:

```{r}
expand<-moanin:::expand_contrast
cns<-expand(moaninObject,contr("K-C",moaninObject)[,1])
colnames(cns)<-gsub("splines::ns(Timepoint, df = 4)","basis",colnames(cns),fixed=TRUE)
cns
```

If we call this matrix $C\in R^{d \times p}$, then our null hypothesis is
$$\theta=C\beta = 0$$
where $\beta$ is the underlying coefficient in our linear model $E(y)=X\beta$, $\theta$ is a $d$-dimensional vector, and $d$ is related to the degrees of freedom of our spline (i.e. how many coefficients in our design matrix $X$ correspond to coefficients in $f_\ell$). 

Currently the package does this implicitly in the code to calculate the fit $\beta_0$ under the null (which then defines the residuals under the null in the calculation of the test-statistics, see below). But the way it is done is based on the assumption that the design matrix $X$ has a coefficient per group in a repeated fashion, which it does if you use the default formula. But this creates problems if there are other variables in the formula separate from the grouping variable. 

I created a function which scans the column names of `basis_matrix` and tries to figure out which ones correspond to level $\ell$ so as to create this matrix $C$ shown above. Then I updated the calculation of  $\beta_0$ to be 

$$\beta_0=(X'X)^{-1}XY-(X'X)^{-1}C'\left( C(X'X)^{-1}C'\right)^{-1}C(X'X)^{-1}XY=\hat\beta - W\hat\beta$$
where
$$W=(X'X)^{-1}C'\left( C(X'X)^{-1}C'\right)^{-1}C$$

and $\hat{\beta}$ is the estimate of $\beta$ under the full (unconstrained model). I have confirmed on the test data in the package that this is giving the same estimates of the $\beta_0$ as the current implementation in the simple grouping situation that the current implementation assumes. 


The code that scans the columns to create $C$ is a weak link, by the way, and probably not robust!


<!-- (currently I have a naive implementation that literally inverts these equations, and I get the same solution as the old code, but I will update it to use Moore-Penrose inverses, etc.) -->

# Test statistics and Implementation details

Both the current and the new implementation provide two test statistics, "fstat" and "lrt" to consider for testing the null hypothesis. 

These are done by the `calculateStat` function, which is an internal function. `calculateStat` takes as input the residuals under the null model (`resNull`) and the residuals under the full model (`resFull`), given as $G x n$ matrices (i.e. each row is the residuals for a gene model). In both implementations, the F-statistic version of the test multiplies the degrees of freedom later when getting the p-values.

## New implementation
1. F-statistic: 
$$F=\frac{(RSS_{null}-RSS_{full})/d}{RSS_{full}/(n-p)} \sim F_{d,n-p}$$
2. Likelihood Ratio statistic:
$$\lambda=2 (\ell(\hat\beta,\hat\sigma)-\ell(\hat\beta_0,\hat\sigma_0))=n\log\frac{RSS_{null}}{RSS_{full}}\sim \chi^2_{d}$$

where $\ell(\beta,\sigma)$ is the log-likelihood of the normal regression model, $\hat\beta, \hat\sigma$ refer to the MLE estimates under the full model, $\hat\beta_0, \hat\sigma_0$ under the null model and for both models, using their respective estimates $\hat\beta$ or $\hat\beta_0$,
$$RSS(\hat\beta)=||Y-X\hat\beta||^2$$
$$\hat\sigma^2=\frac{1}{n}||Y-X\hat\beta||^2$$

This means that $F=\frac{n-p}{d}(e^{n\lambda}-1)$. We would note that some people approach the LRT by calculating it assuming that $\sigma^2$ is known, getting
$$\tilde{\lambda}=\frac{(RSS_{null}-RSS_{full})}{\sigma^2} $$ and then estimate $\sigma^2$ as $RSS_{full}/(n-p)$. This would get us directly back to the $F$ statistic above, up to the constant $d$. In the above calculation, I am not doing this. Instead, I am maximizing the likelihood with $\sigma$ unknown which gives us instead the difference in the *logarithm* of RSS. 

The code in the new implementation is:

```{r eval=FALSE, echo=TRUE}
if(!all(dim(resNull)==dim(resFull))) 
    stop("resNull and resFull must be of equal dimensions")
n<-ncol(resNull)
ss0 <- matrixStats::rowSums2(resNull^2) #RSSNull
ss1 <- matrixStats::rowSums2(resFull^2) #RSSFull
if(type=="lrt") 
    return(n*(log(ss0)-log(ss1)))
# Note that F-statistic returned below is intentionally missing the 
# df terms, which are done later when calculating the p-values 
# (this is for simplicity so don't have to pass df to this function.)
if(type=="ftest") 
    return((ss0 - ss1)/(ss1))
```


## Current implementation

For the lrt, the sum of the squared residuals (RSS) appears to be calculated per group in the grouping variable (getting values $RSS_{null,\ell}$ and $RSS_{full,\ell}$ for a level $\ell$). And the returned statistic is:
$$\sum_\ell \sum_{i=1}^{n_\ell} n_\ell (RSS_{null,\ell}-RSS_{full,\ell})/RSS_{full,\ell}$$

The full code for this is:

```{r eval=FALSE, echo=TRUE}
stat<-0
# ng_labels is a factor that gives the groups in the contrast
for(g in levels(ng_labels)){
    whKeep <- which(ng_labels == g)
    sub_resNull <- resNull[,whKeep]
    sub_resFull <- resFull[,whKeep]
    
    # Somehow the two lines above don't return the same object depending on
    # the dimension of resNull and resFull, so need to distinguish the case
    # where there is only one observation in data.
    if(is.null(dim(sub_resNull))){
        ss0 <- sum(sub_resNull^2)
        ss1 <- sum(sub_resFull^2)
        n <- length(sub_resNull)
    }else{
        ss0 <- matrixStats::rowSums2(sub_resNull^2)
        ss1 <- matrixStats::rowSums2(sub_resFull^2)
        n <- ncol(sub_resNull)
    }
    stat <- stat + n * (ss0 - ss1)/(ss1)
}
```

For the f-test, the statistic appears to be $$n(RSS_{null}-RSS_{full})/RSS_{full}.$$

The code is given by:

```{r eval=FALSE, echo=TRUE}
stat<-0
ss0 <- matrixStats::rowSums2(resNull^2)
ss1 <- matrixStats::rowSums2(resFull^2)
n <- ncol(resNull)
stat <- stat + n * (ss0 - ss1)/(ss1)
```

## Comparison

Both of the definitions in the new implementation for the test-statistics appear to be different from the current implementations. However,  in our comparisons on epicon data, the f-statistic p-values are actually the same in both implementations, and only the LRT p-values changed between the implementations. 

For the f-statistic, the difference between the two implementations appears to be a constant factor of $n$, so it might be that the calculations of the degrees of freedom, which are taken in later code before taking the p-value, correct for this. 

The current LRT seems to actually be more of the average of the F-statistic calculated within each group rather than the LRT and I'm not sure where that comes from.


